---
title: "Assignment: Data mining and statistical tests\n"
author: "Victor Salvat"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: false
    number_sections: true
---

# Load data

```{r load-data}
nonuse <- read.csv("data/test_data/NonUse.csv", sep = ",", header = TRUE)
head(nonuse)
summary(nonuse)
```

This file contains measures of upper extremity non-use in stroke patients. The variables are:

    PANU: Proximal Arm Non Use
    SANU: Shoulder Antepulsion Non Use
    EENU: Elbow Extension Non Use
    
In your notebook, make a clear description of the experiment, the variables and the data. Do not forget to give the units of each variable, and the range of possible values.

| name | description | unit | range |
|:--|:--------|-|--|
| PANU | Proximal Arm Non Use | % | 0-100 |
| SANU | Shoulder Antepulsion Non Use | % | 0-100 |
| EENU | Elbow Extension Non Use | % | 0-100 |


## Corelation analysis
Below are some questions to guide your analysis.

- Is PANU correlated with SANU and/or EENU?

```{r correlation-sanu}
cor.test(nonuse$PANU, nonuse$EENU, method = "spearman")
```
```{r correlation-eenu}
cor.test(nonuse$PANU, nonuse$SANU, method = "spearman")
```


